`gosset`: An R package for analysis and synthesis of ranking data in agricultural experimentation

### 2022-06-23 inst/paper/paper.Rmd
paper updates

::: frontmatter
::: keyword
**Keywords**: Bradley-Terry, Plackett-Luce, data science, on-farm trials, participatory-research, tricot
:::
:::

@@

| **Nr.** | **Code metadata description**                                   | **Please fill in this column**                             |
|:-----------------|:--------------------------|:-------------------------|
| C1      | Current code version                                            | 0.4.003                                                    |
| C2      | Permanent link to code/repository used for this code version    | <https://github.com/AgrDataSci/gosset>                     |
| C3      | Code Ocean compute capsule                                      |                                                            |
| C4      | Legal Code License                                              | MIT                                                        |
| C5      | Code versioning system used                                     | Git                                                        |
| C6      | Software code languages, tools, and services used               | R                                                          |
| C7      | Compilation requirements, operating environments & dependencies |                                                            |
| C8      | If available Link to developer documentation/manual             | <https://agrdatasci.github.io/gosset/>                     |
| C9      | Support email for questions                                     | [kaue.desousa@inn.no](mailto:kaue.desousa@inn.no){.email} |

@@

Participatory experimentation approaches have been increasingly applied in agricultural research [@deRoo2019farm]. While collecting data in ranking format is uncommon in general
agricultural research settings, it is often collected in participatory experiments [@Coe2002AnalyzingRA]. Recently developed approaches for on-farm experimentation,
such as the tricot methodology are based on the collection of data in ranking format [@vanetten_beza_2019]. On the other hand, newly proposed approaches for synthesis of crop variety evaluation data largely depend on the analysis of ranking data [@brown2020data].
The analysis of ranking data requires the use of appropriate statistical models such as Plackett-Luce [@luce_individual_1959; @Plackett] or Bradley-Terry [@bradley1952rank]. Functionality for fitting Bradley-Terry and Plackett-Luce models are available in R with the packages BradleyTerry2 and PlackettLuce respectively [@BradleyTerry2; @Turner2020]. However, extended functionality was required for the entire data science workflow, which usually includes: (1) Data preparation and cleaning, (2) modelling and validation, and (3) results presentation. For (1) gosset provides functions for converting and preparing data into ranking or pairwise format required by the packages PlackettLuce and BradleyTerry2
respectively. For (2), gosset provides functions for model selection and validation using cross-validation. In the case of (3), enhanced functionality for plotting model results is provided by the gosset package.

@@

The R package gosset provides functionality supporting the analysis workflows in agricultural experimentation, especially with rank-based approaches. The package is available in The Comprehensive R Archive Network (CRAN) and can be installed by executing `install.packages("gosset")`. The package is named in honor to William Sealy Gosset, known by the pen name ‘Student’, a pioneer of modern statistics in small sample experimental design and analysis [@Ziliak_2019].

@@

* `plot` provides an alternative method to the default plot function of the PlackettLuce package.  
* `regret` computes the regret [@Loomes_1982;  ].  
* `reliability` computes the reliability of crop varieties, the probability to outperform the reference variety [@eskridge_1992].  
* `worth_bar`  creates a bar plot of the estimated worth values for each evaluated variety.  
* `worth_map` creates a heatmap plot of the estimated worth values for all varieties considering each of the
evaluated traits.

@@

To demonstrate the functionality of the gosset package, we use the `nicabean` dataset, which was generated with decentralized on-farm trials of common bean (*Phaseolus vulgaris* L.) varieties in Nicaragua over five seasons (between 2015 and 2016). Following the tricot  approach [@vanetten_beza_2019], farmers were asked to test in their farms three varieties of common bean. The varieties were randomly assigned as incomplete blocks of size three (out of 10 varieties). The farmers assessed which of the three varieties had the best and worst performance in eight traits (vigor, architecture, resistance to pests, resistance to diseases, tolerance to drought, yield, marketability, taste). The farmers also provided their overall appreciation about the varieties, i.e., which variety had the best and the worst performance based on the overall performance considering all the traits. To analyze the data, we use the Plackett-Luce model implemented in the R package *PlackettLuce* [@Turner2020].

The `nicabean` is a list with two data frames: i) `trial`, contains the trial data with farmers' evaluations, ranked from 1 to 3, with 1 being the higher ranked variety and 3 the lowest ranked variety for the given trait and incomplete block; ii) `covar`, contains the covariates associated to the on-farm trial plots and farmers. This example will require the packages PlackettLuce [@Turner2020], climatrends [@climatrends], nasapower [@nasapower] and ggplot2 [@ggplot2].

@@

To start the analysis of the data, we transform the rankings into Plackett-Luce rankings using the function `rank_numeric()`. We run iteratively over the traits adding the rankings to a list called `R`. Since the varieties are ranked in an ascending order, with 1 being the higher ranked and 3 the lower ranked, we use the argument `asceding = TRUE` to indicate which order should be used.

@@

Then, using the function `kendallTau()` we can assess the correlation between rankings. This approach can be support at least two decision making processes by, for example i) indicating the traits

@@

@@

For each trait, we fit a Plackett-Luce model using the function `PlackettLuce` from the package of the same name. We use ``lapply`` with the list of rankings and the `PlackettLuce` function as input parameters.

@@

The `worth_map` function can be used to visually assess variety performance based on different traits. The values represented in a worth_map are worth estimates in the log scale.

>>>>

::: frontmatter
::: keyword
**Keywords**: Bradley-Terry, Plackett-Luce, data science, on-farm trials, participatory-research, tricot approach
:::
:::

@@

| **Nr.** | **Code metadata description**                                   | **Please fill in this column**                             |
|:-----------------|:--------------------------|:-------------------------|
| C1      | Current code version                                            | 0.4                                                    |
| C2      | Permanent link to code/repository used for this code version    | <https://github.com/AgrDataSci/gosset>                     |
| C3      | Code Ocean compute capsule                                      |                                                            |
| C4      | Legal Code License                                              | MIT                                                        |
| C5      | Code versioning system used                                     | Git                                                        |
| C6      | Software code languages, tools, and services used               | R                                                          |
| C7      | Compilation requirements, operating environments & dependencies |                                                            |
| C8      | If available Link to developer documentation/manual             | <https://agrdatasci.github.io/gosset/>                     |
| C9      | Support email for questions                                     | [kaue.desousa@inn.no](mailto:kaue.desousa@inn.no){.email} |

@@

Participatory experimentation approaches have been increasingly applied in agricultural research [@deRoo2019farm]. While collecting data in ranking format is uncommon in general agricultural research settings, it is often collected in participatory experiments [@Coe2002AnalyzingRA]. Recently developed approaches for on-farm experimentation, such as the tricot methodology are based on the collection of data in ranking format [@vanetten_beza_2019]. On the other hand, newly proposed approaches for synthesis of crop variety evaluation data largely depend on the analysis of ranking data [@brown2020data].

The analysis of ranking data requires the use of appropriate statistical models such as Plackett-Luce [@luce_individual_1959; @Plackett] or Bradley-Terry [@bradley1952rank]. Functionality for fitting Bradley-Terry and Plackett-Luce models are available in R with the packages BradleyTerry2 and PlackettLuce respectively [@BradleyTerry2; @Turner2020]. However, extended functionality was required for the entire data science workflow, which usually includes: (1) Data preparation and cleaning, (2) modelling and validation, and (3) results presentation. For (1) gosset provides functions for converting and preparing data into ranking or pairwise format required by the packages PlackettLuce and BradleyTerry2 respectively. For (2), gosset provides functions for model selection and validation using cross-validation. In the case of (3), enhanced functionality for plotting model results is provided by the gosset package.

@@

The R package gosset provides functionality supporting the analysis workflows in agricultural experimentation, especially with rank-based approaches. The package is available in The Comprehensive R Archive Network (CRAN) and can be installed by executing `install.packages("gosset")`. The package is named in honour to William Sealy Gosset, known by the pen name ‘Student’, a pioneer of modern statistics in small sample experimental design and analysis [@Ziliak_2019].

@@

* `compare` measure the agreement between two methods [@MartinBland1986].
* `plot` provides an alternative method to the default plot function of the PlackettLuce package.  
* `regret` computes the regret coefficients, the loss under the worst possible outcome. A common heuristic in risk assessment strategy [@Loomes_1982].  
* `reliability` computes the probability of a set of items to outperform a reference item. A common heuristic in plant breeding [@eskridge_1992]. 
* `worth_bar`  creates a bar plot of the estimated worth values for each evaluated item.  
* `worth_map` creates a heatmap plot of the estimated worth values for all items considering each of the
evaluated traits.

@@

To demonstrate the functionality of the gosset package, we use the `nicabean` dataset, which was generated with decentralized on-farm trials of common bean (*Phaseolus vulgaris* L.) varieties in Nicaragua over five seasons (between 2015 and 2016). Following the tricot  approach [@vanetten_beza_2019], farmers were asked to test in their farms three varieties of common bean. The varieties were randomly assigned as incomplete blocks of size three (out of 10 varieties). The farmers assessed which of the three varieties had the best and worst performance in eight traits (vigour, architecture, resistance to pests, resistance to diseases, tolerance to drought, yield, marketability, taste). The farmers also provided their overall appreciation about the varieties, i.e., which variety had the best and the worst performance based on the overall performance considering all the traits. To analyse the data, we use the Plackett-Luce model implemented in the R package *PlackettLuce* [@Turner2020].

The `nicabean` is a list with two data frames. The first, `trial`, contains the trial data with farmers' evaluations, ranked from 1 to 3, with 1 being the higher ranked variety and 3 the lowest ranked variety for the given trait and incomplete block. The rankings in this dataset were previously transformed from tricot rankings (where participants indicate best and worst) to ordinal rankings using the function `rank_tricot()`. The second data frame, `covar`, contains the covariates associated to the on-farm trial plots and farmers. This example will require the packages PlackettLuce [@Turner2020], climatrends [@climatrends], nasapower [@nasapower] and ggplot2 [@ggplot2].

@@

To start the analysis of the data, we transform the ordinal rankings into Plackett-Luce rankings using the function `rank_numeric()`. We run iteratively over the traits adding the rankings to a list called `R`. Since the varieties are ranked in an ascending order, with 1 being the higher ranked and 3 the lower ranked, we use the argument `asceding = TRUE` to indicate which order should be used.

@@

Then, using the function `kendallTau()` we can assess the Kendall tau ($\tau$) coefficient [@Kendall1938]. This approach can be used, for example, to assess the drivers of farmers choices or to prioritize traits to be tested in a next stage of tricot trials (e.g. a lite version of tricot with less questions). We use the overall appreciation as the reference trait, and compare the Kendall tau with the other 8 traits.

@@

The kendall correlation shows that farmers prioritized the traits yield ($\tau$ = 0.749), taste ($\tau$ = 0.653) and marketability ($\tau$ = 0.639) when assessing overall appreciation.

@@

Then, for each trait, we fit a Plackett-Luce model using the function `PlackettLuce()` from the package of the same name. This will allow us to continue the analysis of the results using the other functions in the package gosset.

@@

The `worth_map()` function can be used to visually assess item performance based on different characteristics. The values represented in a worth_map are log-worth estimates. From the breeder or product developer perspective the function `worth_map()` offers a visualization tool to help in identifying item performance based on different characteristics.
